Statistically-estimated tree composition for the northeastern United States \\ at the time of Euro-American settlement

### 2016-04-03 paper.tex
adding updated arxiv materials

%
\begin{abstract}
We present a gridded 8 km-resolution data product of the estimated
composition of tree taxa at the time of Euro-American settlement of
the northeastern United States and the statistical methodology used
to produce the product from trees recorded by land surveyors. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height for 22 tree taxa, generally at the genus
level. The data come from settlement-era public survey records that
are transcribed and then aggregated spatially, giving count data.
The domain is divided into two regions, eastern (Maine to Ohio) and
midwestern (Indiana to Minnesota). Public Land Survey point data in
the midwestern region (ca. 0.8-km resolution) are aggregated to a
regular 8 km grid, while data in the eastern region, from Town Proprietor
Surveys, are aggregated at the township level in irregularly-shaped
local administrative units. The product is based on a Bayesian statistical
model fit to the count data that estimates composition on a regular
8 km grid across the entire domain. The statistical model is designed
to handle data from both the regular grid and the irregularly-shaped
townships and allows us to estimate composition at locations with
no data and to smooth over noise caused by limited counts in locations
with data. Critically, the model also allows us to quantify uncertainty
in our composition estimates, making the product suitable for applications
employing data assimilation. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale Euro-American settlement. In addition
to specific regional questions, the data product can also serve as
a baseline against which to investigate how forests and ecosystems
change after intensive settlement. The data product is being made
available at the NIS data portal as version 0.4.

@@

Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
data and processing steps are available \cite{goring2015composition};
key elements include the use of only corner points, the use of only
the two closest trees at each corner point, spatially-varying correction
factors for sampling effort, and a standardized taxonomy table. The
lower Midwest (Illinois, Indiana) data were purchased from the Indiana
State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
Illinois) and processed using similar steps as for the upper Midwest
data. Digitization of the Illinois and Indiana data is still underway,
so many grid cells contained no data at the time the statistical model
was fit. Note that the number of trees per grid cell varies depending
on the number of survey points in a cell, with an average of 124 trees
per cell. The gridded data at the 8 km resolution for the midwest
subdomain are available through the NIS data portal \textcolor{red}{as
DOI:10.5072/FK2/e9c144ab83af70e29b8d6647aaae58f5} \textcolor{red}{\citep{Gori:etal:data:2016}}.
The TPS data were compiled by C.V. Cogbill from a myriad of archival
sources representing land division surveys conducted in connection
with local settlement\textcolor{red}{{} and are available through the
NIS data portal as DOI:10.5072/FK2/4d44aebe71dbf65290d8ce1bde03598b
and DOI:10.5072/FK2/88e167487b22d5f657e7ce7e033772e1 \citep{Cogb:dataNE:2016,Cogb:dataOH:2016}}.

@@

%
\begin{table}
\caption{Predictive ability based on several predictive metric criteria for
the CAR and SPDE spatial models when holding out 95\% of entire cells
of data in Minnesota. Smaller values are better.}

@@

%
\begin{table}
\caption{Predictive ability based on several predictive metric criteria for
the CAR and SPDE spatial models when holding out 80\% of entire cells
of data in Minnesota. Smaller values are better.}

@@

%
\begin{table}
\caption{Predictive ability based on several predictive metric criteria for
the CAR and SPDE spatial models when holding out 5\% of trees. Smaller
values are better.}

@@

The data product is publicly available at the \textcolor{red}{NIS
Data Portal as DOI:10.5072/FK2/dfc5e9748c8df4c3c275eab0f42d9744} under
the CC BY 4.0 license as version 0.4 as of January 2016 \citep{paci:etal:data:2016}.
The product is in the form of a netCDF-4 file, with dimensions x-coordinate,
y-coordinate, and MCMC iteration. There is one variable per taxon.
In addition, dynamic visualizations of the product using the Shiny
tool are available at \href{https://www3.nd.edu/~paleolab/paleonproject/maps}{https://www3.nd.edu/$\sim$paleolab/paleonproject/maps}.
The PalEON Project (in particular the first author) will continue
to maintain this product, releasing new versions as additional data
in Illinois, Indiana and Ohio are digitized. Note that digitization
of data from Illinois and Indiana is ongoing, and digitization of
additional data from Ohio is planned as well. As a result, at some
point we expect to have complete data for the midwestern half of the
domain.

>>>>

\begin{abstract}
We present a gridded 8 km-resolution data product of the estimated
composition of tree taxa at the time of Euro-American settlement of
the northeastern United States and the statistical methodology used
to produce the product from trees recorded by land surveyors. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height for 22 tree taxa, generally at the genus
level. The data come from settlement-era public survey records that
are transcribed and then aggregated spatially, giving count data.
The domain is divided into two regions, eastern (Maine to Ohio) and
midwestern (Indiana to Minnesota). Public Land Survey point data in
the midwestern region (ca. 0.8-km resolution) are aggregated to a
regular 8 km grid, while data in the eastern region, from Town Proprietor
Surveys, are aggregated at the township level in irregularly-shaped
local administrative units. The product is based on a Bayesian statistical
model fit to the count data that estimates composition on a regular
8 km grid across the entire domain. The statistical model is designed
to handle data from both the regular grid and the irregularly-shaped
townships and allows us to estimate composition at locations with
no data and to smooth over noise caused by limited counts in locations
with data. Critically, the model also allows us to quantify uncertainty
in our composition estimates, making the product suitable for applications
employing data assimilation. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale Euro-American settlement. In addition
to specific regional questions, the data product can also serve as
a baseline against which to investigate how forests and ecosystems
change after intensive settlement. The data product is being made
available at the NIS data portal as version 1.0.

@@

Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
data and processing steps are available \cite{goring2015composition};
key elements include the use of only corner points, the use of only
the two closest trees at each corner point, spatially-varying correction
factors for sampling effort, and a standardized taxonomy table. The
lower Midwest (Illinois, Indiana) data were purchased from the Indiana
State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
Illinois) and processed using similar steps as for the upper Midwest
data. Digitization of the Illinois and Indiana data is still underway,
so many grid cells contained no data at the time the statistical model
was fit. Note that the number of trees per grid cell varies depending
on the number of survey points in a cell, with an average of 124 trees
per cell. The gridded data at the 8 km resolution for the midwest
subdomain are available through the NIS data portal \textcolor{red}{\citep{Gori:etal:data:2016}}.
The TPS data were compiled by C.V. Cogbill from a myriad of archival
sources representing land division surveys conducted in connection
with local settlement\textcolor{red}{{} and are available through the
NIS data portal \citep{Cogb:dataNE:2016,Cogb:dataOH:2016}}.

@@

\begin{table}
\caption{Predictive ability based on several predictive metric criteria for
the CAR and SPDE spatial models when holding out 95\% of entire cells
of data in Minnesota. Smaller values are better.}

@@

\begin{table}
\caption{Predictive ability based on several predictive metric criteria for
the CAR and SPDE spatial models when holding out 80\% of entire cells
of data in Minnesota. Smaller values are better.}

@@

\begin{table}
\caption{Predictive ability based on several predictive metric criteria for
the CAR and SPDE spatial models when holding out 5\% of trees. Smaller
values are better.}

@@

The data product is publicly available at the \textcolor{red}{NIS
Data Portal }under the CC BY 4.0 license as version 1.0 as of January
2016 \citep{paci:etal:data:2016}. The product is in the form of a
netCDF-4 file, with dimensions x-coordinate, y-coordinate, and MCMC
iteration. There is one variable per taxon. In addition, dynamic visualizations
of the product using the Shiny tool are available at \href{https://www3.nd.edu/~paleolab/paleonproject/maps}{https://www3.nd.edu/$\sim$paleolab/paleonproject/maps}.
The PalEON Project (in particular the first author) will continue
to maintain this product, releasing new versions as additional data
in Illinois, Indiana and Ohio are digitized. Note that digitization
of data from Illinois and Indiana is ongoing, and digitization of
additional data from Ohio is planned as well. As a result, at some
point we expect to have complete data for the midwestern half of the
domain.
